---
title: "HW R"
Author: "Jahanara Sonia"
date: "3/12/2023"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

 
At first loading all the packages required for this Assignment.

```{r cars}
library (tidyverse)
library(tidyr)
library(dplyr)
library(ggplot2)
summary(cars)
```

Then I loaded the fang_et_al.txt and snp_positions.txt file where the first row will contain the variable names (by the arguement header=TRUE).
```{r cars}
fang_et_al <- read.delim("./fang_et_al_genotypes.txt", header=TRUE)
snp_positions <- read.delim("./snp_position.txt", header = TRUE)
summary(cars)
```

#PART 1: DATA INSPECTION

##Inspection of Fang_et_al_genotypes.txt

 1. Data type is inspected using data frame and vector function.
 2. Then head, tail, dimension and structure are inspected.
 3. View funtion is used to see the data frame.
 
###(Review: only data frames are shown in results, but not the structure and dimension.)
```{r cars}
is.data.frame(fang_et_al)
is.vector(fang_et_al)
head(fang_et_al) 
tail(fang_et_al) 
dim(fang_et_al) 
str(fang_et_al) 
names(fang_et_al) 
sapply(fang_et_al, class) 
View(fang_et_al)
summary(cars)
```

##Inspection of snp_positions.txt
 Same functions as fanf_et_al.txt are used to inspect snp_positions.txt file.

```{r cars}
is.data.frame(snp_positions)
is.vector(snp_positions)
head(snp_positions) 
tail(snp_positions) 
dim(snp_positions) 
str(snp_positions) 
names(snp_positions) 
View(snp_positions)
summary(cars)
```
#DATA PROCESSING

##Snp_postion.txt

  1. Here I have created a new data frame by filtering SNP_ID, Chromosome and Position columns from snp_position.txt and named it new_snp_positions 
  2. Checked the diemnsion(number of columns and rows) of new_snp_positions
  3. Then I exported the data frame to a file naming new_snp_positions.

```{r cars}
new_snp_positions <- snp_positions[,c("SNP_ID", "Chromosome", "Position")] 
dim(new_snp_positions)
write.table(new_snp_positions, './new_snp_positions.txt')
summary(cars)
```

 
##From fang_et_al Creating and Transposing maize files with maize genotypes:
  1.To create maize genotype
  2. Created a data frame by filtering third column "Group" using key words ZMMIL,ZMMLR and ZMMMR and named it maize
  3. Inspected the dimesnsion of the new object maize.
  4. Exported the data frame to a file naming maize.txt
  5. Transposed maize and created a maize_transposed object
  6.  Changed the Sample_ID to SNP_ID

```{r cars}
maize <- fang_et_al %>% filter(Group == "ZMMIL" | Group == "ZMMLR" | Group == "ZMMMR")
dim(maize)
write.table(maize, "./maize.txt")
maize_transposed <- t(maize[,-c(1:3)]) %>% as.data.frame()
maize_transposed$SNP_ID <- rownames(maize_transposed)
view(maize_transposed)
write.table(maize_transposed, "./maize_transposed.txt")
summary(cars)
```

##From fang_et_al Creating and Transposing teosinte files:
 1. Repeated the same steps as maize files to create and transpose teosinte genotype files

```{r cars}
teosinte <- fang_et_al %>% filter(Group == "ZMPBA" | Group == "ZMPIL" | Group == "ZMPJA")
dim(teosinte)
write.table(teosinte, "./teosinte.txt")
teosinte_transposed <- t(teosinte[,-c(1:3)]) %>% as.data.frame()
teosinte_transposed$SNP_ID <- rownames(teosinte_transposed)
view(teosinte_transposed)
write.table(teosinte_transposed, "./teosinte_transposed.txt")
summary(cars)
```

##Joining Maize and Teosinte file with new_snp_positions file:

```{r cars}
maize_joined <- merge(new_snp_positions,maize_transposed,by = 'SNP_ID')
colnames(maize_joined)[4:ncol(maize_joined)] <- as.character(maize$Sample_ID)
maize_joined[,-c(1:3)] <- lapply(maize_joined[,-c(1:3)],as.character)
unique(maize_joined$Chromosome)
write.table(maize_joined, "./maize_joined.txt")
view(maize_joined)

teosinte_joined <- merge(new_snp_positions,teosinte_transposed,by = 'SNP_ID')
colnames(teosinte_joined)[4:ncol(teosinte_joined)] <- as.character(teosinte$Sample_ID)
teosinte_joined[,-c(1:3)] <- lapply(teosinte_joined[,-c(1:3)],as.character)
unique(teosinte_joined$Chromosome)
write.table(teosinte_joined, "./teosinte_joined.txt")
view(teosinte_joined)
summary(cars)
```

##Extracting Maize Chromosome Files: 

   1. To extract maize chromosome files loop is used to create multiple files of chromosomes extracted from maize joined files.
   2. Then the files are arranged in ascending (using subset function) and descending order, as well as missing genotypes are changed to ?mark 
    
```{r cars}
install(gtools)
library(gtools) #install gtools

for (chr in 1:10) {
  maize.chr <- subset(maize_joined,Chromosome == chr) %>% mixedsort(as.character(maize_joined$Position)) #use gtools to make the position as increasing value 
  maize.chr[maize.chr == '?/?'] <- '?' 
  write.table(maize.chr, file = paste("./Maize/maize_chr", chr,"_Ascd.txt",sep = "")) 
  maize.chr_d <- subset(maize_joined,Chromosome == chr) %>% mixedsort(as.character(maize_joined$Position, decreasing = TRUE)) #use gtools to make the position as decreasing value
  maize.chr_d[maize.chr_d == '?/?'] <- '-'
  write.table(maize.chr_d, file = paste("./Maize/maize_chr",chr,"_Desc.txt",sep = ""))
}
summary(cars)
```

##Extracting Teosinte Chromosome Files:

1. Repeated the same steps as maize files to extract teosinte chromosome files

```{r cars}
for (chr in 1:10) {
  teosinte.chr <- subset(teosinte_joined,Chromosome == chr) %>% mixedsort(as.character(teosinte_joined$Position)) #use gtools to make the position as increasing value 
  teosinte.chr[teosinte.chr == '?/?'] <- '?' 
  write.table(teosinte.chr, file = paste("./Teosinte/teosinte_chr", chr,"_Ascd.txt",sep = "")) 
  teosinte.chr_d <- subset(teosinte_joined,Chromosome == chr) %>% mixedsort(as.character(teosinte_joined$Position, decreasing = TRUE)) #use gtools to make the position as decreasing value
  teosinte.chr_d[teosinte.chr_d == '?/?'] <- '-'
  write.table(teosinte.chr_d, file = paste("./Teosinte/teosinte_chr",chr,"_Desc.txt",sep = ""))
}
summary(cars)
```

#Visualization

##1.SNPs Per Chromosome
###1a.Plotting SNP distribution per chromosome

  1. The original SNP file (snp_positions) was filtered by position and the filtering was piped to ggplot function to create a bar graph (using geom_bar function) of total counts of SNPs per chromosome.
  2. X-axis labeled as CHR and the y-axis labeled as Number of SNPs.
  
```{r cars}
snp_positions %>% filter(Position < 10e1000) %>% 
  ggplot(aes(as.double(Chromosome))) + geom_bar(fill = 'blue', color = 'green') + geom_text(stat = 'count', aes(label = ..count..), vjust = -1) +
scale_x_continuous(breaks = 1:10) + theme_replace() + ggtitle("SNPs count/CHR") + ylab('Number of SNPs') + xlab('CHR') 
ggsave('./Visualization/SNPs_number.png')
```

###1b. Plotting SNP distribution across the chromosome

  1. To plot SNP distribution across the chromosome, the snp_positions file was filtered by position was Piped to create histogram SNP distribution across chromosome and fill the histogram with green color
  2. To plot each chromosome facet Wrap function is used and the position was specified as less than infinity.
  3. X-axis labeled as Position and the Y-axis labeled as SNP density. 

```{r cars}
snp_positions %>% filter(Position < 10e1000) %>% 
  ggplot(aes(as.double(Position)/1000000)) + geom_histogram(aes(y = ..density..), color = 'green', fill = "green", alpha = 0.4, bins = 20) + geom_density(aes(as.double(Position)/1000000), color = "blue") + facet_wrap(~ as.double(Chromosome), scales = "free") + theme_replace() + ggtitle("Across Chromosomes distribution") +
  xlab('Position (Mb)') + ylab('SNP density')
ggsave(paste0("./Visualization/SNPs_distribution.png"))

```

##Comparing SNP distribution between maize and teosinte individuals
   The SNP distribution between maize and teosinte chromosome is same.
   
```{r cars}
maize_joined %>% 
  filter(Position < 10e1000) %>% ggplot(aes(as.double(Chromosome))) + geom_bar(fill = 'blue', color = 'green') + 
  geom_text(stat = 'count', aes(label = ..count..), vjust = -1) + scale_x_continuous(breaks = 1:10) + theme_replace() +
  ggtitle("SNPs count/CHR") + ylab('Number of SNPs') + xlab('CHR') 
ggsave('./Visualization/SNPs_number_maize.png')
```
```{r cars}
teosinte_joined %>% filter(Position < 10e1000) %>%  ggplot(aes(as.double(Chromosome))) + geom_bar(fill = 'blue', color = 'green') + 
  geom_text(stat = 'count', aes(label = ..count..), vjust = -1) + scale_x_continuous(breaks = 1:10) + theme_replace() +
  ggtitle("SNPs count/CHR") + ylab('Number of SNPs') + xlab('CHR') 
ggsave('./Visualization/SNPs_number_teosinte.png')
```

##2.Missing data and amount of heterozygosity

###2a. Proportion of homozygous and Hets by sample and Plotting


```{r cars}

mutate_genotypes <- 
  fang_et_al %>% select(-JG_OTU, -Group) %>% pivot_longer(!Sample_ID) %>% 
  mutate(Allele = ifelse(value %in% c('C/C', 'G/G', 'A/A', 'T/T'), 'Homozygote', ifelse(value == '?/?', 'NA', 'Heterozygote')))  

mutate_genotypes %>% group_by(Sample_ID) %>%  count(Allele) %>% 
  ggplot(aes(fill = Allele, y = n, x = Sample_ID)) +
  geom_bar(position = "fill", stat = "identity") +
  ggtitle("Proportion of Homozygotes, Heterozygotes and Missing Data by Sample") +
  ylab('Proportion') +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
ggsave('./Visualization/Allelic_proportion_by_sample.png')

summary(cars)
```

##2b. Proportion of homozygous and Hets by Group and Plotting

```{r cars}

mutate_groups <- 
  fang_et_al %>% select(-Sample_ID, -JG_OTU) %>%   
  pivot_longer(!Group) %>% mutate(Allele = ifelse(value %in% c('C/C', 'G/G', 'A/A', 'T/T'), 'Homozygote', ifelse(value == '?/?', 'NA', 'Heterozygote')))  

mutate_groups %>% group_by(Group) %>%  count(Allele) %>% 
  ggplot(aes(fill = Allele, y = n, x = Group)) +
  geom_bar(position = "fill", stat = "identity") +
  ggtitle("Proportion of Homozygotes, Heterozygotes and Missing Data by Group") +
  ylab('Proportion') 
ggsave('./Visualization/Allelic_Proportion_by_group.jpg')

summary(cars)
```

##3 Own_Plot:Proportion of Heterozygous Allele
Here I have plotted proportion of each Heterozygous Allele in different genotype group.

```{r cars}
mutate_groups %>% filter(Allele == "Heterozygote") %>% group_by(Group) %>%  count(value) %>% 
  ggplot(aes(fill = value, y = n, x = Group)) +
  geom_bar(position = "fill", stat = "identity") +
  ggtitle("Proportion Heterozygote sites in each group") +
  ylab('Proportion') 
ggsave('./Visualization/Heterozygous_distribution.png')
```
## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
